A Statistical Dynamical Model to Predict Extreme Events and Anomalous Features in Shallow Water Waves with Abrupt Depth Change

### 2018-11-27 paper1_tKdV/tKdV_theory.tex
NM

\cite{bolles2018anomalous}

The model parameters $\left(E_{0},L_{0},\Lambda\right)$ in [\ref{eq:tKdV_nondim}]
can be directly linked to the basic scales from the physical problem.
The important parameters that characterize the experiments
include: $\epsilon=\frac{a}{H_{0}}$ the wave amplitude $a$ to (upstream) water
depth $H_{0}$ ratio; $\delta=\frac{H_{0}}{\lambda_{c}}$ the water
depth to wavelength $\lambda_{c}$ ratio; and $D_{0}=\frac{d}{H_{0}}$ 
the depth ratio with upstream value $d=H_{0}$
and downstream value $d<H_{0}$. 
By comparing the characteristic physical scales, the normalized TKdV equation [\ref{eq:tKdV_nondim}] can be
linked directly to the above experimental parameters via
\begin{equation}
L_{0}=6^{\frac{1}{3}}\left(M\epsilon^{\frac{1}{2}}\delta^{-1}\right),\;E_{0}=\frac{27}{2}\gamma^{-2}\left(M\epsilon^{\frac{1}{2}}\delta^{-1}\right),\label{eq:params}
\end{equation}
where $M$ defines the computational domain size $M\lambda_{c}$
as $M$-multiple of the characteristic wavelength $\lambda_{c}$,
and $\gamma=\frac{U}{a}$ normalizes the total
energy of the state variable $u_{\Lambda}$ to one.
\todo[inline]{What is the variable $U$? Have we defined it?}

Consider a spatial discretization with $J=2\Lambda+1$ grid points, so that the smallest
resolved scale is comparable to the characteristic wavelength
\[
\Delta x=\frac{2\pi M\lambda_{c}}{J}\lesssim\lambda_{c}\:\Rightarrow\:M=\frac{J}{2\pi}\sim5,\;J=32.
\]
In the practical numerical simulations, we select $M=5$ and
$\gamma$ varies in the range $\left[0.5,1\right]$. Using the reference
experimental measurements \cite{bolles2018anomalous}, $\epsilon\in\left[0.0024,0.024\right],\delta\sim0.22$,
and $D_{0}$ changes from 1 to 0.24 before and after the depth change.
The reference values for the model scales can be estimated in the
range $L_{0}\in\left[2,6\right]$ and $E_{0}\in\left[50,200\right]$.
These are the values we will test in the direct numerical simulations.
See details about the derivation from scale analysis in \textcolor{blue}{\emph{SI Appendix, A}}.

@@

Since the TKdV equation satisfies the Liouville property, the equilibrium
invariant measure can be described by an equilibrium statistical formulism
\cite{abramov2003hamiltonian,majda2006nonlinear,bajars2013weakly}
using a Gibbs measure with the conserved energy $\mathcal{E}_{\Lambda}$ and
Hamiltonian $\mathcal{H}_{\Lambda}$. The equilibrium invariant measure is dictated
by the conservation laws in the TKdV equation. In the case with fixed
total energy $E_{0}$, this is the \emph{mixed Gibbs measure} in the
truncated model with microcanonical energy and canonical Hamiltonian
ensembles \cite{abramov2003hamiltonian}
\begin{equation}
\mathcal{G}_{\theta}^{\pm}\left(u_{\Lambda}^{\pm};E_{0}\right)=C_{\theta}^{\pm}\exp\left(-\theta^{\pm}\mathcal{H}\left(u_{\Lambda}^{\pm}\right)\right)\delta\left(\mathcal{E}\left(u_{\Lambda}^{\pm}\right)-E_{0}\right),\label{eq:gibbs_mixed}
\end{equation}
with $\theta$ representing the ``inverse temperature''. The distinct
statistics in the upstream and downstream waves can be controlled
by the value of $\theta$. Negative temperature, $\theta^{\pm}<0$, is the appropriate regime to predict the
experiments as shown below. In the incoming flow field, the inverse
temperature $\theta^{-}$ is chosen so that $\mathcal{G}_{\theta}^{-}$
has Gaussian statistics. Using the above invariant measures [\ref{eq:gibbs_mixed}],
the expectation of any functional $F\left(u\right)$ can be computed
based on the Gibbs measure 
\[
\left\langle F\right\rangle _{\mathcal{G}_{\theta}}\equiv\int F\left(u\right)\mathcal{G}_{\theta}\left(u\right)du.
\]
The value of $\theta$ in the invariant measure is specified from
$\left\langle H_{\Lambda}\right\rangle _{\mathcal{G}_{\theta}}$ \cite{abramov2003hamiltonian,bajars2013weakly}.
The invariant measure also predicts an equilibrium energy spectrum
without running the TKdV equation directly. On the other hand, the
time autocorrelation and transient statistics about the state variable
$u_{\Lambda}$ cannot be recovered from the statistical theory.

\subsection{Statistical matching condition in invariant measures before and after the abrupt depth change}

The Gibbs measures $\mathcal{G}_{\theta}^{\pm}$ are defined based
on the different inverse temperatures $\theta^{\pm}$ on the two sides
of the solutions
\[
\begin{aligned}\mu_{t}^{-}\left(u_{\Lambda}^{-};D_{-}\right), & \quad u_{\Lambda}\mid_{t=T_{\mathrm{ADC}-}}=u_{0},\;t<T_{\mathrm{ADC}};\\
\mu_{t}^{+}\left(u_{\Lambda}^{+};D_{+}\right), & \quad u_{\Lambda}\mid_{t=T_{\mathrm{ADC}+}}=u_{0},\;t>T_{\mathrm{ADC}},
\end{aligned}
\]
where $u_{0}$ represents the deterministic matching condition between
the incoming and outgoing waves. The two distributions, $\mu_{t}^{-},\mu_{t}^{+}$
should also be matched at the depth change location $T_{\mathrm{ADC}}$,
so that,
\[
\mu_{t=T_{\mathrm{ADC}}}^{-}\left(u_{\Lambda}\right)=\mu_{t=T_{\mathrm{ADC}}}^{+}\left(u_{\Lambda}\right).
\]
In matching the flow statistics before and after the abrupt depth
change, we first use the conservation of the deterministic Hamiltonian
$H_{\Lambda}^{+}$ after the depth change. Then assuming ergodicity
\cite{abramov2003hamiltonian,majda2006nonlinear}, the statistical
expectation for the Hamiltonian $\left\langle H_{\Lambda}^{+}\right\rangle $
is conserved in time after the depth change at $t=T_{\mathrm{ADC}}$
and should stay in the same value as the system approaches equilibrium
as $t\rightarrow\infty$. The final statistical matching condition
to get the outgoing flow statistics with parameter $\theta^{+}$ can
be found by
\begin{equation}
\left\langle H_{\Lambda}^{+}\right\rangle _{\mathcal{G}_{\theta}^{+}}=\left\langle H_{\Lambda}^{+}\right\rangle _{\mathcal{G}_{\theta}^{-}},\label{eq:matching}
\end{equation}
with the outgoing flow Hamiltonian $H_{\Lambda}^{+}$ and the Gibbs
measures $\mathcal{G}_{\theta}^{\pm}$ before and after the abrupt
depth change.

@@

The incoming flow is always characterized by a near-Gaussian distribution
in the wave displacement. It is found that a physically consistent
Gibbs measure should take negative values in the inverse temperature
parameter $\theta<0$, where a proper distribution function and a
decaying energy spectrum are generated (see \cite{bajars2013weakly}
and \textcolor{blue}{\emph{SI Appendix, B.1}} for the explicit simulation results).
The upstream Gibbs measure $\mathcal{G}_{\theta}^{-}$ with $D_{-}=1$
displays a wide parameter regime in $\left(\theta^{-},E_{0}\right)$
with near-Gaussian statistics. In the left panel of Figure \ref{fig:matching}
(a), the inflow skewness $\kappa_3^{-}$ varies only slightly 
with changing values of $E_{0}$ and $\theta^{-}$. The incoming
flow PDF then can be determined by picking the proper parameter value
$\theta^{-}$ in the near Gaussian regime with small skewness. In
contrast, the downstream Gibbs measure $\mathcal{G}_{\theta}^{+}$
with $D_{+}=0.24$ shown in the right panel of Figure \ref{fig:matching}
(a) generates much larger skewness $\kappa_3^{+}$ 
as the absolute value of $\theta^{+}$ and the total energy level
$E_{0}$ increases. The solid lines in Figure \ref{fig:matching}
(c) offer a further confirmation of the transition from near-Gaussian
statistics with small $\kappa_3^{-}$ to a strongly skewed distribution
$\kappa_3^{+}$ after the depth change.

>>>>

The model parameters $\left(E_{0},L_{0},\Lambda\right)$ in [\ref{eq:tKdV_nondim}]
can be directly linked to the basic scales from the physical problem.
The important parameters that characterize the experiments \cite{bolles2018anomalous}
include: $\epsilon=\frac{a}{H_{0}}$ the wave amplitude $a$ to (upstream) water
depth $H_{0}$ ratio; $\delta=\frac{H_{0}}{\lambda_{c}}$ the water
depth to wavelength $\lambda_{c}$ ratio; and $D_{0}=\frac{d}{H_{0}}$ 
the depth ratio with upstream value $d=H_{0}$
and downstream value $d<H_{0}$. 
By comparing the characteristic physical scales, the normalized TKdV equation [\ref{eq:tKdV_nondim}] can be
linked to these experimental parameters via
\begin{equation}
L_{0}=6^{\frac{1}{3}}\left(M\epsilon^{\frac{1}{2}}\delta^{-1}\right),\;E_{0}=\frac{27}{2}\gamma^{-2}\left(M\epsilon^{\frac{1}{2}}\delta^{-1}\right),\label{eq:params}
\end{equation}
where $M$ defines the computational domain size $M\lambda_{c}$
as $M$-multiple of the characteristic wavelength $\lambda_{c}$,
and $\gamma=\frac{U}{a}$ normalizes the total
energy of the state variable $u_{\Lambda}$ to one.
\todo[inline]{What is the variable $U$? Have we defined it?}

Consider a spatial discretization with $J=2\Lambda+1$ grid points, so that the smallest
resolved scale is comparable to the characteristic wavelength
\[
\Delta x=\frac{2\pi M\lambda_{c}}{J}\lesssim\lambda_{c}\:\Rightarrow\:M=\frac{J}{2\pi}\sim5,\;J=32.
\]
In the practical numerical simulations, we select $M=5$ and
let $\gamma$ vary in the range $\left[0.5,1\right]$. Using reference values from the experiments, 
\cite{bolles2018anomalous}, we find $\epsilon\in\left[0.0024,0.024\right], \delta\sim0.22$,
and $D_{0}$ changes from 1 to 0.24 at the ADC.
These values yield the following the following ranges for the model parameters: $L_{0}\in\left[2,6\right]$ and $E_{0}\in\left[50,200\right]$.
These are the values we will test in the direct numerical simulations.
See details about the derivation from scale analysis in \textcolor{blue}{\emph{SI Appendix, A}}.

@@

Since the TKdV equation satisfies the Liouville property, the equilibrium
invariant measure can be described by a statistical formalism
\cite{abramov2003hamiltonian,majda2006nonlinear,bajars2013weakly}
based on a Gibbs measure with the conserved energy $\mathcal{E}_{\Lambda}$ and
Hamiltonian $\mathcal{H}_{\Lambda}$. The equilibrium invariant measure is dictated
by the conservation laws in the TKdV equation. In the case of fixed energy $E_{0}$, 
this is the \emph{mixed Gibbs measure} with microcanonical energy and canonical Hamiltonian
ensembles \cite{abramov2003hamiltonian}
\begin{equation}
\mathcal{G}_{\theta}^{\pm}\left(u_{\Lambda}^{\pm};E_{0}\right)=C_{\theta}^{\pm}\exp\left(-\theta^{\pm}\mathcal{H}\left(u_{\Lambda}^{\pm}\right)\right)\delta\left(\mathcal{E}\left(u_{\Lambda}^{\pm}\right)-E_{0}\right),\label{eq:gibbs_mixed}
\end{equation}
where $\theta$ represents the ``inverse temperature''. The distinct
statistics in the upstream and downstream waves can be controlled $\theta$. 
As shown below, we find that negative temperature, $\theta^{\pm}<0$, is the appropriate regime to predict the
experiments. In the incoming wave field, $\theta^{-}$ is chosen so that $\mathcal{G}_{\theta}^{-}$
has nearly Gaussian statistics. Using the above invariant measures [\ref{eq:gibbs_mixed}],
the expectation of any functional $F\left(u\right)$ can be computed as
\[
\left\langle F\right\rangle _{\mathcal{G}_{\theta}}\equiv\int F\left(u\right)\mathcal{G}_{\theta}\left(u\right)du.
\]
The value of $\theta$ in the invariant measure is specified from
$\left\langle H_{\Lambda}\right\rangle _{\mathcal{G}_{\theta}}$ \cite{abramov2003hamiltonian,bajars2013weakly}.
\todo[inline]{I do not quite understand the above sentence.}
In addition to producing equilibrium PDFs of $u_{\Lambda}$, the invariant measure can be used to predict the equilibrium energy spectrum without the direct simulation of TKdV. Direct simulation, however, is required to recover transient statistics of $u_{\Lambda}$ and time autocorrelations.
%The invariant measure also predicts an equilibrium energy spectrum without running the TKdV equation directly. On the other hand, the time autocorrelation and transient statistics about the state variable $u_{\Lambda}$ cannot be recovered from the statistical theory.

\subsection{Statistical matching condition of the invariant measures before and after the abrupt depth change}

The Gibbs measures $\mathcal{G}_{\theta}^{\pm}$ are defined based
on the different inverse temperatures $\theta^{\pm}$ on the two sides
of the solutions
\[
\begin{aligned}\mu_{t}^{-}\left(u_{\Lambda}^{-};D_{-}\right), & \quad u_{\Lambda}\mid_{t=T_{\mathrm{ADC}-}}=u_{0},\;t<T_{\mathrm{ADC}};\\
\mu_{t}^{+}\left(u_{\Lambda}^{+};D_{+}\right), & \quad u_{\Lambda}\mid_{t=T_{\mathrm{ADC}+}}=u_{0},\;t>T_{\mathrm{ADC}},
\end{aligned}
\]
where $u_{0}$ represents the deterministic matching condition between
the incoming and outgoing waves. The two distributions, $\mu_{t}^{-},\mu_{t}^{+}$
should be matched at $T_{\mathrm{ADC}}$ giving
\[
\mu_{t=T_{\mathrm{ADC}}}^{-}\left(u_{\Lambda}\right)=\mu_{t=T_{\mathrm{ADC}}}^{+}\left(u_{\Lambda}\right).
\]
In matching the flow statistics before and after the abrupt depth
change, we first use the conservation of the deterministic Hamiltonian
$H_{\Lambda}^{+}$ after the depth change. Then assuming ergodicity
\cite{abramov2003hamiltonian,majda2006nonlinear}, the statistical
expectation for the Hamiltonian $\left\langle H_{\Lambda}^{+}\right\rangle $
is conserved in time after the depth change at $t=T_{\mathrm{ADC}}$
and should stay in the same value as the system approaches equilibrium
as $t\rightarrow\infty$. The final statistical matching condition
to get the outgoing flow statistics with parameter $\theta^{+}$ can
be found by
\begin{equation}
\left\langle H_{\Lambda}^{+}\right\rangle _{\mathcal{G}_{\theta}^{+}}=\left\langle H_{\Lambda}^{+}\right\rangle _{\mathcal{G}_{\theta}^{-}},\label{eq:matching}
\end{equation}
with the outgoing flow Hamiltonian $H_{\Lambda}^{+}$ and the Gibbs
measures $\mathcal{G}_{\theta}^{\pm}$ before and after the abrupt
depth change.

@@

For the parameters explored in \cite{bolles2018anomalous}, the incoming wave field 
is always characterized by a near-Gaussian distribution
of the surface displacement. It is found that a physically consistent
Gibbs measure should take negative values in the inverse temperature
parameter $\theta<0$, where a proper distribution function and a
decaying energy spectrum are generated (see \cite{bajars2013weakly}
and \textcolor{blue}{\emph{SI Appendix, B.1}} for the explicit simulation results).
The upstream Gibbs measure $\mathcal{G}_{\theta}^{-}$ with $D_{-}=1$
displays a wide parameter regime in $\left(\theta^{-},E_{0}\right)$
with near-Gaussian statistics. In the left panel of Figure \ref{fig:matching}
(a), the inflow skewness $\kappa_3^{-}$ varies only slightly 
with changing values of $E_{0}$ and $\theta^{-}$. The incoming
flow PDF then can be determined by picking the proper parameter value
$\theta^{-}$ in the near Gaussian regime with small skewness. In
contrast, the downstream Gibbs measure $\mathcal{G}_{\theta}^{+}$
with $D_{+}=0.24$ shown in the right panel of Figure \ref{fig:matching}
(a) generates much larger skewness $\kappa_3^{+}$ 
as the absolute value of $\theta^{+}$ and the total energy level
$E_{0}$ increases. The solid lines in Figure \ref{fig:matching}
(c) offer a further confirmation of the transition from near-Gaussian
statistics with small $\kappa_3^{-}$ to a strongly skewed distribution
$\kappa_3^{+}$ after the depth change.
